---
title: "00 explore and create cannabis dataset"
author: "Ryan Schubert"
date: "July 7, 2021"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(readxl)
library(dplyr)
library(naniar)
library(tidyr)
library(ggplot2)
"%&%" = function(a,b) paste0(a,b)
```

## read in complete data and subset to relevant variables


```{r}
dir<-"C:\\Users\\rshoo\\OneDrive\\Desktop\\rush\\cannabis use\\"
completeData<-read_xlsx(dir %&% 'IOP_DATA_SET_6-12-2020 with Manual Corrections.xlsx',na = "-")
#write out a csv for faster reading later
#fwrite(completeData,dir %&% 'IOP_DATA_SET_6-12-2020 with Manual Corrections.csv')
#subset to variables of interest
#DAYSABST2,DAYSABST2_POST,DAYSABST2_3MON,DAYSABST2_6MON,DAYSABST2_12MON
#aware that days abstinent actually measures days substance was used
#also outome variables
#PCL_SCORE,PCL[1-20]_DAY14
#basic patient info
#PAT_ID

vars<-c("PAT_ID","DAYSABST2","DAYSABST2_POST","PCL_SCORE","PCL_DAY14")
cannabisData<-completeData %>% select(one_of(vars))
#next engineer the total PCL score for day 14


vis_miss(cannabisData)
fwrite(cannabisData,dir %&% 'IOP_cannabis_outcomes.csv')
# cannabisData<-cannabisData[complete.cases(cannabisData),]
# fwrite(cannabisData,dir %&% 'IOP_cannabis_outcomes_complete_cases.csv')
```

fewer samples in complete data set than I would have expected - should be around 500, but have 480 instead?
double check that I have correct variables

# exploratory

remember that days abstinent actually measures days used in the last two weeks

```{r}
str(cannabisData)
hist(cannabisData$DAYSABST2)
sum(cannabisData$DAYSABST2 > 0)

hist(cannabisData$DAYSABST2_POST)
sum(cannabisData$DAYSABST2_POST > 0)

table(cannabisData$DAYSABST2 > 0,cannabisData$DAYSABST2_POST > 0)

```
most people have used no cannabis in the two weeks prior to the IOP with ~21% having used at all
similarly most do not use cannabis in the final two weeks of the IOP with only ~11% having used at all
In total ~25% of people used cannabis at any time point


a small proportion of people who did not use prior to the IOP went on to use during the IOP
A large proportion of peple who did use prior to the IOP went on to not use during the final two weeks of the IOP

# basic analysis

first I want to know if use at each individual time point is associated with outcome scores and change in scores
Then I want to know if being a user at either time point is associated with outcome scores and change in scores
then I want to know if number of days are associated with outcome scores and change in outcome scores

## First create additional variables

```{r}
cannabisData<-cannabisData %>% mutate(user_pre=if_else(cannabisData$DAYSABST2 > 0,T,F),
                        user_concurrent=if_else(cannabisData$DAYSABST2_POST > 0,T,F),
                        user_either=if_else(cannabisData$DAYSABST2 > 0 | cannabisData$DAYSABST2_POST > 0,T,F),
                        delta_PCL=PCL_SCORE-PCL_DAY14)
hist(cannabisData$delta_PCL)
```

## Check some asusmptions


```{r}
shapiro.test(cannabisData$delta_PCL) # change in score is not normal
shapiro.test(cannabisData$PCL_DAY14) # outcome score is not normal

```


## Next analyze groupings

```{r}



cannabisDataLong<-cannabisData %>% pivot_longer(starts_with("PCL")) %>%
  mutate(name=if_else(name=="PCL_SCORE","Baseline","Post"))

cannabisDataLong %>% filter(complete.cases(cannabisDataLong)) %>%
  group_by(name,user_pre) %>%
  summarise(PCL_outcomes=mean(value)) %>%
  ggplot(aes(x=name,y=PCL_outcomes,colour=user_pre)) +
    geom_line(aes(group=user_pre,colour=user_pre)) +
    geom_point(aes(colour=user_pre)) +
    theme_bw()

cannabisDataLong %>% filter(complete.cases(cannabisDataLong)) %>%
  group_by(name,user_concurrent) %>%
  summarise(PCL_outcomes=mean(value)) %>%
  ggplot(aes(x=name,y=PCL_outcomes,colour=user_concurrent)) +
    geom_line(aes(group=user_concurrent,colour=user_concurrent)) +
    geom_point(aes(colour=user_concurrent)) +
    theme_bw()

cannabisDataLong %>% filter(complete.cases(cannabisDataLong)) %>%
  group_by(name,user_either) %>%
  summarise(PCL_outcomes=mean(value)) %>%
  ggplot(aes(x=name,y=PCL_outcomes,colour=user_either)) +
    geom_line(aes(group=user_either,colour=user_either)) +
    geom_point(aes(colour=user_either)) +
    theme_bw()

ggplot(data=cannabisData,aes(x=PCL_DAY14,y=PCL_SCORE)) +
  geom_point(aes(colour=user_pre)) +
  geom_smooth(aes(group=user_pre,colour=user_pre),method="lm",se=F) +
  theme_bw() +
  geom_abline(slope=1)




```

# model analysis

next run categories with models

```{r}
summary(linear_regression)

library(lme4)
library(lmerTest)
cannabis_interaction_regression1<-lmer(value ~ name*user_pre + (1|PAT_ID),data=cannabisDataLong)
summary(cannabis_interaction_regression1)
cannabis_interaction_regression1b<-lmer(value ~ (name|PAT_ID) + user_pre + name:user_pre +  (1|PAT_ID),data=cannabisDataLong)
anova(cannabis_interaction_regression1,cannabis_interaction_regression1b)
#having a random slope for treatment effect does not lower the AIC, don't expect it to be relevant for other categories 


cannabis_interaction_regression2<-lmer(value ~ name*user_concurrent + (1|PAT_ID),data=cannabisDataLong)
summary(cannabis_interaction_regression2)

cannabis_interaction_regression3<-lmer(value ~ name*user_either + (1|PAT_ID),data=cannabisDataLong)
summary(cannabis_interaction_regression3)


```

can reasonably conclude that there is a significant treatment effect, but no effect or interaction with cannabis use status at any time point

lookup paired tests for multiple treatment groups, anova?


## continuous and ordinal models

```{r}

cannabis_continuous_model1<-lmer(value ~ name*DAYSABST2_POST + (1|PAT_ID),data=cannabisDataLong)
summary(cannabis_continuous_model1)

cannabis_continuous_model2<-lmer(value ~ name*DAYSABST2 + (1|PAT_ID),data=cannabisDataLong)
summary(cannabis_continuous_model2)

cannabisDataLong <-mutate(cannabisDataLong,DAYSABST2_TOTAL = DAYSABST2 + DAYSABST2_POST)


cannabis_continuous_model3<-lmer(value ~ name*DAYSABST2_TOTAL + (1|PAT_ID),data=cannabisDataLong)
summary(cannabis_continuous_model3)

```

# Next lets look at the LMM taking all PCL days into account plus the factors that were accounted for in 

doi: 10.1080/20008198.2021.1888541

age, sex, and cohort type

conditionally independent errors selected based on AIC

included random slopes and intercepts 

```{r}
vars<-c("PAT_ID","COHORT_TYPE","AGE_AT_COHORT_START","gender","DAYSABST2","DAYSABST2_POST","PCL_SCORE","PCL_DAY" %&% 2:14)
cannabisDataExtended<-completeData %>% select(one_of(vars))
cannabisDataExtendedLong<-cannabisDataExtended %>% pivot_longer(starts_with("PCL")) %>%
  mutate(name=factor(name, ordered = TRUE, levels = unique(cannabisDataExtendedLong$name)),
         DAYSABST2_TOTAL=DAYSABST2 + DAYSABST2_POST,
         user_pre=if_else(DAYSABST2 > 0,T,F),
         user_concurrent=if_else(DAYSABST2_POST > 0,T,F),
         user_either=if_else(DAYSABST2_TOTAL > 0,T,F))
  
cannabisDataExtendedLong %>% filter(complete.cases(cannabisDataExtendedLong)) %>%
  group_by(name,user_either) %>%
  summarise(PCL_outcomes=mean(value)) %>%
  ggplot(aes(x=name,y=PCL_outcomes,colour=user_either)) +
    geom_line(aes(group=user_either,colour=user_either)) +
    geom_point(aes(colour=user_either)) +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 45,, hjust=1)) +
  ylim(0,60)

cannabisDataExtendedLong %>% filter(complete.cases(cannabisDataExtendedLong)) %>%
  group_by(name,user_concurrent) %>%
  summarise(PCL_outcomes=mean(value)) %>%
  ggplot(aes(x=name,y=PCL_outcomes,colour=user_concurrent)) +
    geom_line(aes(group=user_concurrent,colour=user_concurrent)) +
    geom_point(aes(colour=user_concurrent)) +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 45,, hjust=1)) +
  ylim(0,60)

cannabisDataExtendedLong %>% filter(complete.cases(cannabisDataExtendedLong)) %>%
  group_by(name,user_pre) %>%
  summarise(PCL_outcomes=mean(value)) %>%
  ggplot(aes(x=name,y=PCL_outcomes,colour=user_pre)) +
    geom_line(aes(group=user_pre,colour=user_pre)) +
    geom_point(aes(colour=user_pre)) +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 45,, hjust=1)) +
  ylim(0,60)


total_use_model<-lmer(value ~ name*DAYSABST2_TOTAL + AGE_AT_COHORT_START + gender + COHORT_TYPE + (1|PAT_ID),data=cannabisDataExtendedLong)
anova(total_use_model)

prior_use_model<-lmer(value ~ name*DAYSABST2 + AGE_AT_COHORT_START + gender + COHORT_TYPE + (1|PAT_ID),data=cannabisDataExtendedLong)
anova(prior_use_model)

concurrent_use_model<-lmer(value ~ name*DAYSABST2_POST + AGE_AT_COHORT_START + gender + COHORT_TYPE + (1|PAT_ID),data=cannabisDataExtendedLong)
#summary(cocurrent_use_model)
anova(concurrent_use_model)
```


